---
title: "Whale Encounters"
author: "S. Wotherspoon"
date: "02/09/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(geosphere)
library(ggplot2)
```

## Data

```{r}
d <- read.csv("Encounters.csv",header=T)
d$Date.start <- as.POSIXct(strptime(paste(d$Date,d$Encounter.start),"%d/%m/%Y %I:%M:%S %p",tz="GMT"))
d$Date.end <- as.POSIXct(strptime(paste(d$Date,d$Encounter.Finish),"%d/%m/%Y %I:%M:%S %p",tz="GMT"))
d$Date <- as.POSIXct(strptime(d$Date,"%d/%m/%Y",tz="GMT"))
```

The ordering of records in the data appears wrong, so we order by date, boat and encounter number, and then check that "same as last" remains correct.
```{r}
d <- d[order(d$Date,d$Boat,d$EncounterID),]
```

Check the ordering of the encounter starts
```{r}
any(sapply(split(d,d[,c("Date","Boat")],drop=TRUE), function(d) is.unsorted(d$Date.start)))
```

The gaps between some encounter times are too great for the "same as last" indicator to be reliable. 
```{r}
k <- which(d$Same.as.last!="N")
d$Date.start[k]-d$Date.start[k-1]
```

Find common groups by comparing group structure and encounter times and positions.
```{r}
k <- 2:nrow(d)
dx <- distGeo(cbind(d$Longitude[k],d$Latitude[k]),
              cbind(d$Longitude[k-1],d$Latitude[k-1]))
dt <- d$Date.start[k]-d$Date.start[k-1]
same <- 
  ## Same boat
  (d$Boat[k]==d$Boat[k-1]) &
  ## Same composition
  (d$Group.Size[k]==d$Group.Size[k-1]) &
  (d$Group.comp...AD[k]==d$Group.comp...AD[k-1]) &
  (d$Group.comp...Calf[k]==d$Group.comp...Calf[k-1]) &
  (d$Cow.calf.pair[k]==d$Cow.calf.pair[k-1]) &
  ## Same encounter time
  dt == 0 &
  ## Within 200m?
  dx < 200
```

Drop encounters with the same group
```{r}
d.u <- d[c(TRUE,!same),]
```

We will only compare distances between encounters from the same boat and date
```{r}
k <- 2:nrow(d.u)
index <- k[
   ## Same boat
  (d.u$Boat[k]==d.u$Boat[k-1]) &
   ## Same date
  (d.u$Date[k]==d.u$Date[k-1])]
## Inter-encounter distances
dist <- distGeo(cbind(d.u$Longitude[index],d.u$Latitude[index]),
                cbind(d.u$Longitude[index-1],d.u$Latitude[index-1]))
```


AS an initial test, we compare the distance between encounters of the same type (cow/calf vs adults). 

First we calculate the differnce in means distance for encounters of the same type and encounters of different types.
```{r}
cowcalf <- d.u$Cow.calf.pair
same <- cowcalf[index]==cowcalf[index-1]
diff.obs <- mean(dist[!same]) - mean(dist[same])
```

Permute the cow/calf vs adult labels and test the difference in mean distances
```{r}
n <- 1000
diffs <- double(n)
for(k in 1:n) {
  cowcalf <- sample(d.u$Cow.calf.pair)
  same <- cowcalf[index]==cowcalf[index-1]
  diffs[k] <- mean(dist[!same]) - mean(dist[same])  
}
hist(diffs)
abline(v=diff.obs)
mean(diff.obs > diffs)
```


